\guidelinesubsubsection{Recommendations}

When conducting any kind of empirical study involving LLMs, researchers \must clearly declare that an LLM was used (see \scope for what we consider relevant research support).
This \should be done in a suitable section of the \paper, for example, in the introduction or research methods section.
For authoring scientific articles, this transparency is, for example, required by the ACM Policy on Authorship: \enq{The use of generative AI tools and technologies to create content is permitted but must be fully disclosed in the Work}~\cite{ACM2023}.
Beyond generic authorship declarations and declarations that LLMs were used as part of the research process, researchers \should report the exact purpose of using an LLM in a study, the tasks it was used to automate, and the expected benefits in the \paper.
% Related paper:
% Artificial intelligence‑assisted academic writing: recommendations for ethical use
% https://doi.org/10.1186/s41077-025-00350-6

\guidelinesubsubsection{Example(s)}

The \emph{ACM Policy on Authorship} suggests to disclose the usage of GenAI tools in the acknowledgments section of the \paper, for example: \emph{``ChatGPT was utilized to generate sections of this Work, including text, tables, graphs, code, data, citations''}~\cite{ACM2023}. 
Similarly, the acknowledgments section could also be used to disclose GenAI usage for other aspects of the research, if not explicitly described in other sections.
The ACM policy further suggests: \emph{``If you are uncertain about the need to disclose the use of a particular tool, err on the side of caution, and include a disclosure in the acknowledgments section of the Work''}~\cite{ACM2023}.
For double-blind review, during which the acknowledgments section is usually hidden, researchers can add a temporary ``AI Disclosure'' section in the same place the acknowledgments section would appear. 
An example of an LLM disclosure beyond writing support can be found in a recent paper by \citeauthor{DBLP:conf/re/LubosFTGMEL24}~\cite{DBLP:conf/re/LubosFTGMEL24}, in which they write in the methodology section:

\begin{quote}
\small
\it
``We conducted an LLM-based evaluation of requirements utilizing the Llama 2 language model with 70 billion parameters, fine-tuned to complete chat responses...''
\end{quote}

\guidelinesubsubsection{Benefits}

Transparency in the use of LLMs helps other researchers understand the context and scope of the study, facilitating better interpretation and comparison of the results.
Beyond this declaration, we recommend researchers to be explicit about the LLM version they used (see Section \modelversion) and the LLM's exact role (see Section \toolarchitecture).

\guidelinesubsubsection{Challenges}

In general, we expect following our recommendations to be straightforward.
One challenge might be authors' reluctance to disclose LLM usage for valid use cases, because they fear that AI-generated content makes reviewers think that the authors' work is less original.
In fact, there is evidence suggesting that AI disclosure can negatively affect trust in authors~\cite{SCHILKE2025104405}.
However, the \emph{ACM Policy on Authorship} is very clear in that any use of GenAI tools to create content \must be disclosed.
Before the introduction of LLMs, human proof-reading was possible and allowed and was not required to be declared.
Our guidelines focus on research support beyond proof-reading and writing support (see \scope).
However, over time, the threshold of what must be declared and what not will most likely evolve.

\guidelinesubsubsection{Study Types}

Researchers \must follow this guideline for all study types.

%This is probably the longest reviewer advice because of the complexity of disclosure failures.
\guidelinesubsubsection{Advice for Reviewers}

\added{The most common problem with disclosure will be incompleteness or vagueness around how, exactly, the LLM was used. The paper will say `we used LLM X to help with task Y' and the reviewer will wonder `help how?' In these cases, reviewers should simply ask for more details. Unless the reviewer suspects that these ``more details'' will reveal some serious problem(s), such clarification requests are minor revisions.}

\added{Many papers will contain exculpatory statements to the effect of `LLM X was used to copyedit and polish the text of this paper'. Using an LLM as a copyeditor becomes problematic when human authors do not or cannot take responsibility for the resulting text. If a reviewer finds evidence that conclusively demonstrates that the authors did not carefully review text generated by an LLM (e.g. the paper includes a ridiculous AI-generated image,%\footnote{e.g. \url{https://arstechnica.com/science/2024/02/scientists-aghast-at-bizarre-ai-rat-with-huge-genitals-in-peer-reviewed-article/}}
text like ``As a Large Language Model, I...''; or AI-generated charts or statistics) the paper should be rejected.} 

\added{An even more problematic scenario is an author who \textit{cannot} take responsibility for AI-generated or -altered text, because, for example, they lack sufficient language skills. If a reviewer suspects this is the case, they should discuss it with their editor or program chair. \textit{Never} accuse authors of misconduct without clear evidence.}

\added{Similarly, if one suspects that an LLM was used to support a study in an undisclosed way, the reviewer should discuss the issue with their editor or program chair. In the absence of conclusive evidence, these cases must be handled delicately without rushing to accuse authors of wrongdoing. If the evidence \textit{is} conclusive, reviewers should question the degree to which the undisclosed LLM use affects the study's contribution---large effect suggests rejecting the paper; no effect suggests leniency. For example, the 18th author used ChatGPT to find the right word (``exculpatory)'' for the sentence above. Failing to disclose this is irrelevant because it has no effect on the study's contribution. In contrast, any undisclosed use of an LLM to generate ostensibly empirical data, or statical analyses (e.g. p-values hallucinated by an LLM) would be grounds for immediate rejection.}