When using LLMs for empirical studies in SE, researchers face unique challenges and potential limitations that can influence the validity, reliability, and reproducibility of their findings~\cite{sallou2024breaking}.
It is important to openly discuss these limitations and explain how their impact was mitigated.
All this is relative to the current capabilities of LLMs and the current state of the art in terms of tool architectures.
If and how the performance of LLMs in a particular study context will improve with future model generations or new architectural patterns is beyond what researchers can and should discuss as part of a paper's limitation section.
Nevertheless, risk management and threat mitigation are important tasks during the design of an empirical study.
They should not happen as an afterthought.

\guidelinesubsubsection{Recommendations}
Researchers \must make every effort to present the limitations of their work clearly without defensiveness or obfuscation. These limitations may concern diverse topics including generalizability, internal validity (e.g. data leakage), reliability (i.e. non-determinism), and reproducibility (e.g. resource requirements). 
\added{When deterministic reproducibility is structurally unattainable (e.g., SaaS-based models with opaque versioning), researchers \should adopt trustworthiness criteria from qualitative research to substantiate dependability and confirmability of findings.}

LLM-based studies may involve many kinds of generalization including the following:
\begin{enumerate}
    \item From tested LLMs to others; that is, do the performance characteristics of the LLM(s) studied generalize to LLM(s) not included in the study.
    \item From tested configurations to others; that is, how sensitive are the results to the specific configuration(s) of the LLM under test. 
    \item From research to practice; for instance, just because researchers can get a certain level of code generation performance under ideal conditions does not mean software developers, given the same tools, will get similar results without having the researchers present to show them exactly what to do. 
    \item From a sample of people (e.g. human validators; human participants) to a larger population of people.
    \item From one time period to another; that is, does the same LLM, or other versions of the same LLM, produce similar results at a different point in time?
\end{enumerate}

\paragraph{Internal Validity}

The primary threats to internal validity are:
\begin{itemize}
	\item Data leakage and contamination, including inter-dataset duplication, potentially resulting in a training–evaluation overlaps, leading to overly optimistic evaluation results.
	\item Unintended inclusion of evaluation data in model-improvement pipelines, contributing to data leakage. This is especially relevant for longitudinal studies including LLMs.
	\item Incomplete architecture, prompt, or pipeline reporting, posing hidden confounding factors.
\end{itemize}

Inter-dataset duplication is a prevlanet issue in the SE research community. An example is code-related benchmarks. As transparency on training data is limited for LLMs, researchers \must discuss potential data leakage effects and their impact on results.

\paragraph{Construct Validity}
The primary threats to construct validity are:
\begin{itemize}
	\item Metric–construct mismatch (e.g., BLEU/ROUGE vs. functional correctness). More traditional metrics might not capture the all relevant SE-specific aspects.
	\item Over-reliance on benchmark-specific metrics. Optimizing for a single benchmark might result sin dataset-specific heuristics rather than the intended construct, overstating real-world utility.
	\item Benchmark scope limitations that omit relevant SE properties. Benchmark setups commonly ignore runtime behaviors such as execution-time failures, performance regressions, security implications, code readability, testability, and maintainability, yielding results that may not transfer to realistic development settings.
\end{itemize}

If constructs are based on subjective interpretations e.g., usage of LLMs as Tools for Software Engineering Research, purely automated metrics are insufficient. Researchers \must discuss how they ensured quality of the subjective results, similarly as for qualitative research.

\paragraph{External Validity}
The primary threats to external validity are:
\begin{itemize}
    \item Cross-model transfer limitations. %TODO (Florian): Extend
    \item Tool-architecture specificity. %TODO (Florian): Extend
    \item Limited domain coverage. %TODO (Florian): Extend
    \item Limited participant diversity. %TODO (Florian): Extend
    \item Cross-time instability (model evolution). Performance of proprietary models can change over time, leading to non-generalizable study outcomes~\cite{DBLP:journals/corr/abs-2307-09009, doi:10.1148/radiol.232411}.
\end{itemize}

Generalizability is particularly critical for proprietary and non-deterministic systems whose behaviors is subject to drift. Those limitations and employed mitigations for external validity \must be discussed by the researchers.

\paragraph{Reliability \& Reproducibility}
The primary threats to reliability and reproducibility are:
\begin{itemize}
    \item Non-deterministic outputs. %TODO (Florian): Extend
    \item Infrastructure dependence. %TODO (Florian): Extend
    \item Resource inequality preventing replication. LLM research is resource-intensive and hence remains predominantely in the domain of private companies or well-funded research institutions excluding researchers e.g., from the global south.
\end{itemize}

The non-deterministic nature of LLMs hinders reliable statements based on single LLM executions. Further, reproducibility is constrained by the dependence on infrastructure, model stability and unequal resource access (e.g., global south). Researchers \must discuss the measures to increase reliability and reproducibility.
\added{However, non-deterministic reproducibility is not inherently disqualifying. Qualitative traditions such as ethnography, grounded theory, and action research ensure trustworthiness through \emph{credibility}, \emph{transferability}, \emph{dependability}, and \emph{confirmability}~\cite{guba1981criteria}.
SaaS-based LLM research faces analogous conditions, as providers frequently deprecate model versions without guaranteeing stable behavior.}

\paragraph{Ethical \& Regulatory Boundaries}
The primary concerns for ethical and regulatory matters are:
\begin{itemize}
    \item Use of sensitive or proprietary data. %TODO (Florian): Extend
    \item Jurisdictional obligations. %TODO (Florian): Extend
    \item Implicit model bias. Especially for qualitative research LLM might \enq{reinforce dominant paradigms and biases} and \enq{identify, replicate and reinforce dominant language and patterns}~\cite{jowsey2025reject}.
\end{itemize}

Studies involving sensitive data \must include data governance mechanisms tailored towards LLM environments, compliant with applicable juristical obligations. If applicable, researchers \should discuss how model biases potentially impact the study outcomes and how those biases were evaluated.

\paragraph{Environmental \& Sustainability Constraints}
\begin{itemize}
    \item Energy consumption. With growing model size, the environmental impact of experiments with LLMs increases. Considering the environmental impact in the study design is an ethical responsibility.
    \item Trade-off between repetition and sustainability. Increased reliability through repetition and the aim of cautious energy consumption pose an inherent tension field to navigate during the study design.
\end{itemize}

The study findings \should be discussed in the context of the resource-intensiveness of LLMs to justify the increased resource consumption over traditional approaches.

\paragraph{Mitigation Strategies}
The following mitigation strategies may serve as inspiration to approach some of the before mentioned threats depending on the study scope and feasibility. 

\textbf{Replication Packages} covering prompt and architecture specifications, models outputs, and representative examples for partial replicability. Also accompanied by an implementation using an open model for long-term stability. See %TODO (Florian): Link to Section 5.2-5.4-5.7

\textbf{Human Validation} of subjective constructs, following quality criteria known from qualitative research. See %TODO (Florian): Link to section 5.5

\textbf{Longitudinal Re-Runs} to repeat experiments with LLMs over time, complemented by statistical analyses. See %TODO (Florian): Link to section 5.4?

\textbf{Methodological Trustworthiness Measures}
Researchers \should therefore consider triangulation, reflexivity, audit trails (see \prompts, and peer debriefing as complementary measures when deterministic reproduction is structurally impossible.

\textbf{Triangulation} via multiple models (e.g., proprietary and open models), multiple independent datasets and multiple complementary metrics. See  %TODO (Florian): Link to sections 5.2,5.3,

\textbf{Cost Accounting} through reporting of input/output tokens, token and service costs, or hardware specifications. %TODO (Florian): Link to relevant sections

\textbf{Energy Preservation} by selecting smaller or newer less resource-intensive models and employing techniques such as input/output token reduction, model pruning, quantization, or knowledge distillation~\cite{mitu2024hidden} where feasible. Carbon footprint estimation is desirable, but still difficult.

\textbf{Ethical and Regulatory Considerations} through data governance mechanisms, ethical reviews, and bias assessment procedures. See %TODO (Florian): Link to section 5.7

\textbf{Sensitivity Analysis} through variation of LLM configurations, prompts, architecture decisions, datasets, and if applicable human participant backgrounds. %TODO (Florian): Link to 5.2,5.3,