\section{Introduction}
\label{sec:introduction}


\input{_scope/01_motivation.tex}

\section{Scope of Guidelines}
\label{sec:scope}

\input{_scope/02_scope.tex}

\section{Methodology}
\label{sec:methodology}

This project was initiated at the 2024 meeting of the \emph{International Software Engineering Research Network} (ISERN) in Barcelona, Spain, where the first and last author organized a session on guidelines for empirical studies involving LLMs.
These discussions led to a position paper at \emph{WSESE 2025}~\citep{DBLP:conf/wsese/0001BFB25}, which was presented at the \emph{\nth{2} Copenhagen Symposium on Human-Centered Software Engineering AI} (CHASEAI 2024), forming a workstream to collaboratively refine and extend the preliminary guidelines.
Our process follows an established tradition of developing reporting guidelines through expert collaboration, as exemplified by CONSORT for clinical trials~\citep{Begg1996} and, more recently, TRIPOD-LLM for LLM-based prediction studies in medicine~\citep{Gallifant2025}.

We held bi-weekly meetings, in which we decided on the structure of the study type taxonomy and the guidelines.
During these discussions, we added a new study type (\synthesis), refined and extended the guideline sections (in particular \modelversion, \toolarchitecture, \prompts) and covered \benchmarksmetrics as well as \limitationsmitigations.

Then, at least two co-authors were assigned to each study type and guideline to refine and extend the content, followed by a peer review and refinement phase.
Afterwards, the first author reviewed all sections and discussed potential changes with the topic leads.
A requirement for co-authorship was that each author contributed or reviewed content and, most importantly, read the complete guidelines and confirmed that they stand behind all recommendations, not only their own contributions.

To select illustrative examples for each study type and guideline, co-author pairs independently searched for relevant papers using academic databases.
This initial pool was extended based on suggestions from other co-authors.
Identified papers were assessed by one author and subsequently reviewed by a second author; papers that did not add new insights beyond those already covered were not included.
Multiple review rounds were then carried out on the guidelines as a whole, during which selected papers were cross-checked by additional authors and, where necessary, further examples were added or replaced.
The examples are intended to be illustrative, not the result of a systematic review; the involvement of multiple authors at different stages helped mitigate individual selection bias.

After completing the internal review, we invited external experts in empirical SE to review the study types and guidelines.
During the review process of the \emph{Empirical Software Engineering} journal, the structure and content were improved based on reviewers' suggestions.
We self-assessed our guidelines against the quality dimensions of the \emph{SIGSOFT Empirical Standards}~\citep{ralph2021empiricalstandardssoftwareengineering}; this assessment is reported in the conclusion.


\section{Study Types}
\label{sec:study-types}

Empirical guidelines are needed for SE studies involving LLMs to ensure the validity and reproducibility of their results.
However, such guidelines must be tailored to different study types that each pose unique challenges.
%Therefore, understanding the classification of these studies is essential for developing appropriate guidelines.
Therefore, we created a taxonomy of study types to contextualize our recommendations.
%Note that our \guidelines refer to the \studytypes but not the other way around.
We use the term \emph{study type} to refer to categories of LLM involvement in empirical research, rather than to research methodologies (e.g., experiments, case studies). A single empirical study may involve multiple such study types.
Each study type section starts with a \textbf{description}, followed by \textbf{examples} from the SE research community and beyond. The \textbf{advantages} and \textbf{challenges} of using LLMs are discussed in a summary subsection at the end of this section, synthesizing cross-cutting themes across all study types.
See Table~\ref{tab:study-types} for an overview of study types.

%% This sub-table of contents is organizational overkill. 
%% I think it's  a good additional since the guidlines are already a wall of text.

\begin{table}[tb]
\caption{Overview of study types.}
\label{tab:study-types}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{ID} & \textbf{Section} & \textbf{Topic} \\
\midrule
     & 4.1   & \textbf{\hyperref[sec:llms-as-tools-for-software-engineering-researchers]{LLMs as Tools for Software Engineering Researchers}} \\
S1   & 4.1.1 & \quad \hyperref[sec:llms-as-annotators]{LLMs as Annotators} \\
S2   & 4.1.2 & \quad \hyperref[sec:llms-as-judges]{LLMs as Judges} \\
S3   & 4.1.3 & \quad \hyperref[sec:llms-for-synthesis]{LLMs for Synthesis} \\
S4   & 4.1.4 & \quad \hyperref[sec:llms-as-subjects]{LLMs as Subjects} \\
\midrule
     & 4.2   & \textbf{\hyperref[sec:llms-as-tools-for-software-engineers]{LLMs as Tools for Software Engineers}} \\
S5   & 4.2.1 & \quad \hyperref[sec:studying-llm-usage-in-software-engineering]{Studying LLM Usage in Software Engineering} \\
S6   & 4.2.2 & \quad \hyperref[sec:llms-for-new-software-engineering-tools]{LLMs for New Software Engineering Tools} \\
S7   & 4.2.3 & \quad \hyperref[sec:benchmarking-llms-for-software-engineering-tasks]{Benchmarking LLMs for Software Engineering Tasks} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{LLMs as Tools for Software Engineering Researchers}
\label{sec:llms-as-tools-for-software-engineering-researchers}

\input{_studytypes/01_llms-as-tools-for-software-engineering-researchers.tex}

\input{_studytypes/01-01-llms-as-annotators.tex}

\input{_studytypes/01-02-llms-as-judges.tex}

\input{_studytypes/01-03-llms-for-synthesis.tex}

\input{_studytypes/01-04-llms-as-subjects.tex}

\subsection{LLMs as Tools for Software Engineers}
\label{sec:llms-as-tools-for-software-engineers}

\input{_studytypes/02_llms-as-tools-for-software-engineers.tex}

\input{_studytypes/02-01_studying-llm-usage-in-software-engineering.tex}

\input{_studytypes/02-02_llms-for-new-software-engineering-tools.tex}

\input{_studytypes/02-03_benchmarking-llms-for-software-engineering-tasks.tex}

\subsection{Advantages and Challenges}
\label{sec:advantages-challenges}

\input{_studytypes/01-05-advantages-and-challenges.tex}

\section{Guidelines}
\label{sec:guidelines}

A primary goal of our guidelines is to \emph{enable reproducibility and replicability} of empirical SE studies involving LLMs. As repeating LLM-focused research to verify results lies somewhere between the ACM's definitions of reproducibility (different team, same research artifacts) and repeatability (different team, different research artifacts)~\cite{acm2020artifactreview} due to potential model changes and imperfect research artifacts, we follow \citeauthor{DBLP:journals/corr/abs-2510-25506}~\cite{DBLP:journals/corr/abs-2510-25506} and use the terms interchangeably. While previous guidelines regarding open science and empirical studies still apply, LLM-specific characteristics (e.g. inherent non-determinism~\cite{DBLP:conf/naacl/SongWLL25, YuanNondeterminismLLMInference}, opaque and proprietary models) present additional replicability challenges, which, in turn, demand new guidance.  

Each guideline below begins with a brief \tldr summary, followed by its \textbf{rationale}, \textbf{recommendations}, \textbf{examples}, \textbf{benefits}, \textbf{challenges}, links to the relevant \textbf{study types}.
The \emph{rationale} articulates the underlying principle, i.e., why the guideline matters, while the \emph{recommendations} provide concrete, actionable practices.
Table~\ref{tab:rationale-recommendations} provides a compact overview of this mapping.

The guidelines further contain \textbf{advice for reviewers}.
Broadly, reviewers should use our guidelines to help them interrogate the extent to which a manuscript's authors have done what was practically possible to improve reproducibility. We must neither accept research absent reasonable efforts to improve reproducibility, nor reject research for failing to obtain an impossible goal. We borrow this principle of ``reasonable efforts'' from the SIGSOFT Empirical Standards~\cite{ralph2021empiricalstandardssoftwareengineering}, where it applies to methodological rigor more generally. Of course, papers should acknowledge their limitations, but to determine whether these limitations are reasonable, reviewers should ask ``have the authors done what they could to minimize limitations?'' Our guidelines attempt to capture what ``reasonable efforts'' practically means for LLM-based studies.

To distinguish essential criteria from recommendations, our guidelines use two tiers.
A~\must criterion is a \emph{requirement}. Studies that intend to follow these
guidelines are expected to meet all \must criteria.
A~\should criterion represents a desired practice that strengthens a study's rigor or transparency.
However, there may be valid reasons to deviate in particular circumstances (e.g., resource constraints, inapplicability to a specific study context or type).
Nonetheless, studies that deviate from a \should criterion should briefly justify the
deviation and discuss its potential impact (e.g., on validity or reproducibility).
Table~\ref{tab:guidelines} lists our eight guidelines.
Table~\ref{tab:guideline-matrix} provides an overview of how each guideline applies to each study type.
Cells marked \iconM indicate that the guideline is a \must requirement for the study type and cells marked \iconS indicate that the guideline \should be followed.
Cells marked -- indicate that the guideline is typically not applicable or not feasible for the study type.

The following sections indicate which information we expect researchers to report, and whether it should be in the \paper or \supplementarymaterial. Where a publication venue's page limits hinder reporting all expected elements in the \paper, it is better to report essential information in the \supplementarymaterial than not at all.
The \supplementarymaterial should be published according to the \emph{ACM SIGSOFT Open Science Policies}~\citep{daniel_graziotin_2024_10796477}.

\begin{table}[tb]
\caption{Overview of guidelines.}
\label{tab:guidelines}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{ID} & \textbf{Section} & \textbf{Guideline} \\
\midrule
G1 & \ref*{sec:declare-llm-usage-and-role} & \hyperref[sec:declare-llm-usage-and-role]{Declare LLM Usage and Role} \\
G2 & \ref*{sec:report-model-version-configuration-and-customizations} & \hyperref[sec:report-model-version-configuration-and-customizations]{Report Model Version, Configuration, and Customizations} \\
G3 & \ref*{sec:report-tool-architecture-beyond-models} & \hyperref[sec:report-tool-architecture-beyond-models]{Report Tool Architecture Beyond Models} \\
G4 & \ref*{sec:report-prompts-their-development-and-interaction-logs} & \hyperref[sec:report-prompts-their-development-and-interaction-logs]{Report Prompts, their Development, and Interaction Logs} \\
G5 & \ref*{sec:use-human-validation-for-llm-outputs} & \hyperref[sec:use-human-validation-for-llm-outputs]{Use Human Validation for LLM Outputs} \\
G6 & \ref*{sec:use-an-open-llm-as-a-baseline} & \hyperref[sec:use-an-open-llm-as-a-baseline]{Use an Open LLM as a Baseline} \\
G7 & \ref*{sec:use-suitable-baselines-benchmarks-and-metrics} & \hyperref[sec:use-suitable-baselines-benchmarks-and-metrics]{Use Suitable Baselines, Benchmarks, and Metrics} \\
G8 & \ref*{sec:report-limitations-and-mitigations} & \hyperref[sec:report-limitations-and-mitigations]{Report Limitations and Mitigations} \\
\bottomrule
\end{tabular}
\end{table}

\input{_summary/matrix.tex}

\subsection{Declare LLM Usage and Role (G1)}
\label{sec:declare-llm-usage-and-role}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/01_declare-llm-usage-and-role-tldr.tex}
\end{framed}

\input{_guidelines/01_declare-llm-usage-and-role}

\subsection{Report Model Version, Configuration, and Customizations (G2)}
\label{sec:report-model-version-configuration-and-customizations}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/02_report-model-version-configuration-and-customizations-tldr.tex} 
\end{framed}

\input{_guidelines/02_report-model-version-configuration-and-customizations.tex}

\subsection{Report Tool Architecture Beyond Models (G3)}
\label{sec:report-tool-architecture-beyond-models}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/03_report-tool-architecture-beyond-models-tldr.tex} 
\end{framed}

\input{_guidelines/03_report-tool-architecture-beyond-models.tex}

\subsection{Report Prompts, their Development, and Interaction Logs (G4)}
\label{sec:report-prompts-their-development-and-interaction-logs}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/04_report-prompts-their-development-and-interaction-logs-tldr.tex}
\end{framed}

\input{_guidelines/04_report-prompts-their-development-and-interaction-logs.tex}

\subsection{Use Human Validation for LLM Outputs (G5)}
\label{sec:use-human-validation-for-llm-outputs}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/05_use-human-validation-for-llm-outputs-tldr.tex}  
\end{framed}

\input{_guidelines/05_use-human-validation-for-llm-outputs.tex}

\subsection{Use an Open LLM as a Baseline (G6)}
\label{sec:use-an-open-llm-as-a-baseline}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/06_use-an-open-llm-as-a-baseline-tldr.tex} 
\end{framed}

\input{_guidelines/06_use-an-open-llm-as-a-baseline.tex}

\subsection{Use Suitable Baselines, Benchmarks, and Metrics (G7)}
\label{sec:use-suitable-baselines-benchmarks-and-metrics}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/07_use-baselines-benchmarks-and-metrics-tldr.tex}  
\end{framed}

\input{_guidelines/07_use-baselines-benchmarks-and-metrics.tex}

\subsection{Report Limitations and Mitigations (G8)}
\label{sec:report-limitations-and-mitigations}

\vspace{0.5\baselineskip}
\begin{framed}
\tldr: \input{_tldr/08_report-limitations-and-mitigations-tldr.tex}   
\end{framed}

\input{_guidelines/08_report-limitations-and-mitigations.tex}

\input{_summary/rationale-recommendations.tex}

\section{Conclusion}

LLM non-determinism, opaque training data, and rapidly evolving models
threaten the reproducibility and replicability of empirical SE studies.
This paper presents a taxonomy of seven study types that organizes how
LLMs are used in SE research, together with eight guidelines for
designing and reporting such studies.
Each guideline distinguishes requirements (\must) from recommended
practices (\should) and is contextualized by the study types it
applies to.
Our guidelines recommend that researchers:
\begin{enumerate*}[label=(\arabic*)]
\item declare LLM usage and role;
\item report model versions, configurations, and customizations;
\item document the tool architecture beyond the model;
\item disclose prompts, their development, and, where feasible,
interaction logs;
\item validate LLM outputs with humans;
\item include an open LLM as a baseline;
\item use suitable baselines, benchmarks, and metrics; and
\item articulate limitations and mitigations, including costs, potential
biases, and environmental impact.
\end{enumerate*}
An applicability matrix maps guidelines to study types, and a reporting
checklist (Section~\ref{sec:checklist}) supports authors and reviewers.
The study types and guidelines were derived collaboratively by
22~co-authors, following a process similar to that of guidelines in
other disciplines (see, e.g.,~\cite{Begg1996}).

Following the quality dimensions of the \emph{SIGSOFT Empirical
Standards}~\cite{ralph2021empiricalstandardssoftwareengineering}, we
consider our guidelines
(1)~\emph{comprehensive}, as they cover seven study types that we
identified and were derived from 22 co-authors' collective expertise;
(2)~\emph{useful}, as they provide actionable recommendations with an
applicability matrix and a concrete reporting checklist
(Section~\ref{sec:checklist});
(3)~\emph{well-argued}, as each guideline is motivated by a dedicated
rationale and supported by references to the literature; and
(4)~\emph{integrated with previous work}, as they build on and
complement established empirical SE standards while addressing
LLM-specific challenges.
Adopting these practices will not eliminate LLM non-determinism, but it
should make claims auditable and results easier to replicate and compare
across studies.
Because models and tools evolve, we maintain the guidelines as a living
resource and invite the community to refine and extend them at
\href{https://llm-guidelines.org/}{llm-guidelines.org}.

\section*{Acknowledgments}

Our study types and guidelines are based on discussion sessions with researchers at the \emph{2024 International Software Engineering Research Network} (ISERN) meeting and at the \emph{2nd Copenhagen Symposium on Human-Centered Software Engineering AI} (supported by the Carlsberg Foundation with grant CF24-0693 and the Alfred P. Sloan Foundation with grant G-2024-22586 to Daniel Russo).
LLM-based tools, including ChatGPT (GPT-5) and Claude Code (Opus 4.6), were used interactively for language editing and structural revision suggestions across all sections of this paper. All LLM-generated suggestions were reviewed and revised by the authors.

\begin{appendices}

\input{_summary/checklist.tex}

\end{appendices}