In the short period since the release of ChatGPT in November 2022, large language models (LLMs) have changed the software engineering (SE) research landscape.
Although numerous opportunities exist for LLMs to support SE research and development tasks, solid science needs rigorous empirical evaluations to explore the effectiveness, performance, and robustness of using LLMs to automate different research and development tasks.
LLMs can, for example, support literature reviews, reduce development time, and generate software documentation.
However, it is often unclear how valid, reproducible, and replicable empirical studies involving LLM are.
This uncertainty poses significant challenges for researchers and practitioners who seek to draw reliable conclusions from empirical studies. \added{Recent findings indicate a low reproducibility of SE studies involving LLMs~\cite{DBLP:journals/corr/abs-2510-25506}, raising concerns about the contemporary studies long-term value.}
%\added{If a new research team repeats a published study, either using the original study's data (reproduction) or using the same research method with new data (replication), and gets opposite results, we would be mad not to question the original study's conclusions.} 

Open science practices and documenting a study's setup are necessary to facilitate reproduction and replication~\cite{DBLP:journals/corr/abs-2412-17859}.
Studies involving LLMs may be irreproducible and irreplicable due to not only their inherent non-determinism  causing variability in model performance but also differences in configuration, training data, model architecture, or evaluation metrics.
Slight changes can lead to significantly different results~\cite{DBLP:journals/corr/abs-2412-12509, DBLP:conf/naacl/SongWLL25, DBLP:conf/nips/AgarwalSCCB21, bjarnason2026randomnessagenticevals}.
For replication, researchers need detailed information on the exact study setup, evaluation procedures, benchmarks, and metrics used. 
These issues highlight the need for clear guidelines and best practices for designing and reporting studies involving LLMs.

Although the SE research community has developed guidelines for conducting and reporting specific types of empirical studies such as controlled experiments (e.g.,~\cite{DBLP:books/sp/WohlinRHORW24, DBLP:books/sp/08/SSS2008}), their replications (e.g.,~\cite{DBLP:journals/tse/SantosVOJ21}), or empirical studies in general (e.g., the \emph{ACM SIGSOFT Empirical Standards}~\cite{ralph2021empiricalstandardssoftwareengineering}), there has been no guidance for using LLMs in empirical studies.
LLMs have characteristics that require specific guidelines vis-\'a-vis reproducibility and replicability~\cite{DBLP:conf/wsese/0001BFB25}.
For example, even if we knew the specific version of a commercial LLM used for an empirical study, the reported task performance could still change over time, since commercial models evolve beyond version identifiers~\cite{DBLP:journals/corr/abs-2307-09009}.
Moreover, commercial providers do not guarantee the availability of old model or tool versions indefinitely.
LLM performance also varies widely depending on configured parameters such as temperature.
Therefore, not reporting the parameter settings severely impacts reproducibility.

%% Paul thinks this paragraph should be moved elsewhere. Here, it distracts from the flow of the argument. 
%Evaluating LLMs' performance is complicated by the fact that researchers often cannot tell whether the solution to a certain problem was part of the training data. Even for ``open'' models such as \emph{Mistral, DeepSeek or Llama}, we do not know how they were fine-tuned for specific tasks and what the exact training data was~\cite{Gibney2024}.


Therefore, we aim to provide holistic guidelines for conducting and reporting studies involving LLMs in SE research.
After outlining our \scope, we introduce a taxonomy of \studytypes, then presenting eight \guidelines co-developed by the authors of this article.
For each study type and guideline, we identify relevant research exemplars, both within and outside SE. The most recent version of the guidelines is always available online.\footnote{\url{https://llm-guidelines.org}}
We encourage interested researchers to suggest improvements via a public GitHub repository.\footnote{\url{https://github.com/se-uhd/llm-guidelines-website}}