In this section, we outline the scope of the proposed guidelines.

\scopeparagraph{Software Engineering as our Target Discipline}

We target SE research (conducted in academic or industry contexts) because SE research has unique considerations. For example, while Gallifant et al.'s~\cite{Gallifant2025} guidelines for LLM use in healthcare research include items that apply to SE (e.g., reporting the LLM name and version), they also include items irrelevant to most SE studies (e.g., ``usability of the LLM in the context of current clinical care''). Meanwhile, Gallifant et al.'s guidelines do not include the same level of specificity around reporting exact prompts (e.g. their structure, content, formatting, and dynamic components) needed for SE research. 

Moreover, SE research utilizes a broader range of empirical methods than many other disciplines. \citeauthor{sallou2024breaking} \cite{sallou2024breaking} presented a vision paper on mitigating validity threats in LLM-based SE research. Our work builds on their guidelines in many ways, including providing more detailed advice for a wider variety of research methods, organized into a taxonomy of \studytypes (below).

\scopeparagraph{Focus on Textual Use Cases}

While multi-modal foundational models (which use or generate images, audio, or video) have great potential to support SE research and practice, we begin with textual use cases of LLMs (e.g., in natural language (NL) or programming languages) to keep our initial guidelines tractable. We hope future work can extend these guidelines to additional modalities. 

\scopeparagraph{Focus on Direct Tool or Research Support}

LLMs are already widely used to support several aspects of the overall research process, from proof-reading, spell-checking, and text translation, to data coding and synthesis of literature reviews. 
For engineers, we focus on the use of LLMs to automate SE tasks---artificial intelligence (AI) for software engineering (AI4SE)---(see~\llmsforengineers). %Don't write ``See section \llmsforengineers'' because the sections aren't numbered.
For researchers, we focus on the use of LLMs to automate empirical research tasks such as data collection, processing, or analysis (see~\llmsforresearcher). \reviewText{However, using LLMs to design studies or write reports is beyond our current scope.}

\scopeparagraph{Researchers as our Target Audience}

Our guidelines are intended to help SE researchers design, plan, conduct, and report empirical studies involving LLMs. While the same guidelines can support scholarly peer review of SE studies using LLMs, this aim is secondary. 