\studytypesubsection{LLMs for Synthesis}
\label{sec:llms-for-synthesis}

\studytypeparagraph{Description}

Unlike annotation (see Section \annotators), which focuses on categorizing or labeling individual data points, refers to the process of integrating and interpreting information from multiple sources to generate higher-level insights, identify patterns across datasets, and develop conceptual frameworks or theories. LLMs may be able to support synthesis tasks in SE research by processing and distilling information from qualitative data sources. 

% Synthesis tasks can also mean generating synthetic datasets (e.g., source code, bug-fix pairs, requirements, etc.) that are then used in downstream tasks to train, fine-tune, or evaluate existing models or tools.
% In this case, the synthesis is done primarily using the LLM and its training data; the input is limited to basic instructions and examples.

\studytypeparagraph{Example(s)} 

Published examples of applying LLMs for synthesis in SE remain scarce; however, some recent work in other domains is instructive~\cite{DBLP:journals/ase/BanoHZT24}.
\citeauthor{barros2024largelanguagemodelqualitative}~\cite{barros2024largelanguagemodelqualitative} conducted a systematic mapping study on using LLMs for qualitative research and found that most studies focused on the healthcare and social sciences. In these studies, LLMs supported qualitative methods, such as grounded theory and thematic analysis, by aiding in pattern identification.
%They identified examples in domains such as healthcare and social sciences (see, e.g., \cite{de2024performing,mathis2024inductive}) in which LLMs were used to support qualitative analysis, including grounded theory and thematic analysis.
%Overall, the findings highlight the successful generation of preliminary coding schemes from interview transcripts, later refined by human researchers, along with support for pattern identification.This approach was reported not only to expedite the initial coding process but also to allow researchers to focus more on higher-level analysis and interpretation.
%This particular paper suggests using LLMs to support tasks such as initial coding and theme identification while conservatively reserving interpretative or creative processes for human analysts.
\added{In SE, de Morais Leca et al. \cite{leça2024applicationsimplicationslargelanguage} explored how LLMs have been applied for qualitative data analysis (QDA) and proposed general strategies and guidelines for their application. Ornelas et al. \cite{ornelas2025llm} complemented this perspective by studying the opportunities and limitations of introducing LLM-based support into QDA, and by formulating recommendations for embedding human–AI collaboration across the thematic analysis phases. Building on these, subsequent work has proposed hybrid frameworks combining LLM support with human-led QDA. Rashid et al. \cite{DBLP:journals/corr/abs-2402-01386} designed an LLM-driven multi-agent system that integrates AI with human decision-making to automate qualitative data analysis methods. Their system generated initial codes, developed themes, and summarized text. Similarly, Montes et al. \cite{DBLP:journals/corr/abs-2510-18456} compared the performance of humans and LLMs in coding, theme development, definition, and refinement, creating guidelines for a hybrid-LLM framework.}

% Finally, \citeauthor{DBLP:journals/corr/abs-2506-21138}'s work is an example of using LLMs to create synthetic datasets.
% They present an approach to generate synthetic requirements, showing that they \enq{can match or surpass human-authored requirements for specific classification tasks}~\cite{DBLP:journals/corr/abs-2506-21138}.

\studytypeparagraph{Advantages}

\added{All of the studies in the previous paragraph reported reducing time, cognitive demands, and resources required for qualitative research.} As discussed, qualitative research does not scale well to large samples and using LLMs allows researchers to analyze larger qualitative datasets~\cite{DBLP:journals/ase/BanoHZT24,barros2024largelanguagemodelqualitative, leça2024applicationsimplicationslargelanguage}.
%Bano24 is a position paper. It does not provide any evidence that using LLMs for synthesis improves generalizability, etc. barros2024 found that LLMs improved efficiency and reduced disagreements. No mention of generalizability or scaling. 
LLMs \textit{may} improve consistency and help researchers develop, refine, and define themes~\cite{DBLP:conf/chi/ByunVS23}.
%I removed a bunch of speculation here. We should avoid speculating and avoid repeating other author's speculations. 

\studytypeparagraph{Challenges}

LLMs may misinterpret nuanced qualitative data due to a lack of contextual understanding and struggle to capture contextual nuances ~\cite{bano2023exploringqualitativeresearchusing}. Other studies have echoed similar concerns~\cite{DBLP:journals/ase/BanoHZT24, barros2024largelanguagemodelqualitative, leça2024applicationsimplicationslargelanguage}.\citeauthor{barros2024largelanguagemodelqualitative} emphasize that effective use of LLMs requires structured prompts and careful human oversight. LLMs can neither engage in reflexivity (which is central to qualitative analysis, as it requires researchers to actively interpret data) nor reflect on their own positionality beyond merely extracting patterns~\cite{DBLP:journals/corr/abs-2510-18456}.
%Popular LLMs and LLM-based tools tend to overgeneralize results when summarizing scientific articles, that is, they \enq{produce broader generalizations of scientific results than those in the original.} Compared to human-authored summaries, \enq{LLM summaries were nearly five times more likely to contain broad generalizations}~\cite{peterschineyee2025generalizationbias}.
In addition, LLMs can produce biased results, reinforcing existing prejudices or omitting essential perspectives, making human oversight crucial to ensure accurate interpretation, mitigate biases, and maintain quality control. Their proprietary nature and unknown training data limits transparency around such biases, and their synthesis may depend on both model versions and prompt variations. 

\added{More fundamentally, the nature of LLMs is philosophically and methodologically incongruent with qualitative research. ``GenAI remains simulated intelligence only, based on statistical predictive algorithms without any understanding of the world, or the meaning of the language that constitutes the data being analyzed, or indeed the meaning of the resulting themes produced when simulating qualitative analysis. While GenAI with human involvement might be able to produce something that superficially resembles reflexive qualitative analysis \dots, it cannot be reflexive, because, by definition, reflexive qualitative analysis is an inherently meaning-based technique.''~\cite{jowsey2025reject}} These qualitative researchers are not confused about what an LLM is; it is the SE researchers who are confused about what qualitative research is. 
 

%\par\noindent\hdashrule{\columnwidth}{0.5pt}{2pt 2pt}

% Reverted back to "subject", as participant seems to be primarily used for humans: see  https://dictionary.apa.org/subject and https://dictionary.apa.org/participant
\guidelinesubsubsection{LLMs as Subjects}
\label{sec:llms-as-subjects}

\studytypeparagraph{Description}

In empirical studies, data is collected from participants through methods such as surveys, interviews, or controlled experiments.
LLMs can serve as virtual \emph{participants} by simulating human behavior and interactions. If LLMs can generate responses that approximate those of human participants, they could be valuable for research involving user interactions, collaborative coding environments, and software usability assessments~\cite{ZHAO2025101167}. \added{Furthermore, a key advantage is the automation of tasks previously executed by humans}.

To achieve this, prompt engineering techniques are widely employed; for instance, the \textit{Personas Pattern}~\cite{DBLP:journals/corr/abs-2308-07702} involves tailoring LLM responses to align with predefined profiles or roles that emulate specific user archetypes.

To serve as virtual participants, generated responses should be: 
\begin{enumerate}
    \item indistinguishable from human-produced texts; 
    \item consistent with the attitudes and sociodemographic information of the conditioning context (e.g., LLMs simulating junior developers should exhibit different confidence levels, vocabulary, and concerns than senior engineers);
    \item naturally aligned with the form, tone, and content of the provided context (e.g., responses in an agile stand-up meeting simulation should be concise, task-focused, and aligned with sprint objectives rather than long, formal explanations); 
    \item reflect patterns in relationships between ideas, demographics, and behavior observed in comparable human data (e.g., discussions on software architecture decisions should capture trade-offs typically debated by human developers, such as maintainability versus performance, rather than abstract theoretical arguments)~\cite{DBLP:journals/corr/abs-2209-06899}.
\end{enumerate}

\studytypeparagraph{Example(s)}

\citeauthor{DBLP:journals/ipm/XuSRGPLSH24}~\cite{DBLP:journals/ipm/XuSRGPLSH24} compiled a list of ways LLMs can support social science research, some of which transfer to empirical SE research. For example, LLMs can can emulate human responses and behaviors in simulated interviews and focus groups~\citeauthor{DBLP:journals/ase/GerosaTSS24}~\cite{DBLP:journals/ase/GerosaTSS24}. Similarly, \citeauthor{bano2025doessoftwareengineerlook}~\cite{bano2025doessoftwareengineerlook}, investigated biases in LLM-generated candidate profiles in SE recruitment processes. (They found biases favoring male candidates, lighter skin tones, and slim physiques, particularly for senior roles.) LLMs may be able to simulate end-user feedback and behavior in usability studies, identify usability issues and offering suggestions for improvement based on predefined user personas. 

\studytypeparagraph{Advantages}

Augmenting or replacing human participants with LLM-generated virtual participants would reduce human participant recruitment, which is often time-consuming and costly~\cite{DBLP:conf/vl/Madampe0HO24}. \textit{If} virtual participants' behavior is sufficiently similar to human participants, LLMs could (virtually) access underrepresented and hard-to-reach populations, strengthening generalizability and inclusiveness, impute missing data (see \synthesis), \added{and open up whole new areas of research that are ethically problematic with real humans (e.g. questions that would force a human to relive past trauma do not harm an LLM).} 

\studytypeparagraph{Challenges}

\added{The \textit{if} in the previous paragraph is doing some heavy lifting. There is no evidence that LLMs can accurately simulate the actual behavior of an end-user in a usability study or a software professional completing a survey.  \enq{LLMs do not simulate human psychology and \dots researchers should treat LLMs as useful but fundamentally unreliable tools that need to be validated against human responses for every new application}~\cite{schroeder2025llmspsychology}. In fact, given a ``novel moral vignette'', for example, the veracity of an LLM's reaction ``is impossible to assess without doing further confirmatory work with human participants''~\cite{DBLP:journals/ais/HardingDLL24}. Meanwhile, ``LLMs are likely to both misportray and flatten [i.e. fail to capture the diversity of opinions and experiences within a group] the representations of demographic groups''~\cite{DBLP:journals/corr/abs-2402-01908} so they cannot actually simulate access to underrepresented groups. Identity-based prompts can reduce identities to stereotypes and amplify perceived differences between groups. While creative prompting or other strategies may partially mitigate some of these problems, again, \textit{there is no evidence that LLMs can accurately simulate human participants, such evidence is needed for each specific LLM usage, and gathering this evidence may be quite difficult.}}
