\studytypesubsection{Benchmarking LLMs for Software Engineering Tasks}
\label{sec:benchmarking-llms-for-software-engineering-tasks}

\studytypeparagraph{Description}

Benchmarking is the process of evaluating an LLM's performance on standardized tasks, using standardized metrics, on a standardized dataset. The LLM output is compared to a supposed ground truth from the benchmark dataset. Typical tasks include code generation, code summarization, code completion, and code repair, but also natural language processing tasks, such as anaphora resolution (i.e., the task of identifying the referring expression of a word or phrase occurring earlier in the text). %, interesting for subfields such a Requirements Engineering. 
Metrics may include general metrics for text generation, such as \emph{ROUGE}, \emph{BLEU}, or \emph{METEOR}~\cite{10.1145/3695988}, or task-specific metrics, such as \emph{CodeBLEU} for code generation.  Benchmarking requires high-quality reference datasets.

\studytypeparagraph{Example(s)}

In SE, benchmarking may include evaluating an LLM's ability to produce accurate and reliable outputs for a given input (usually a task description, possibly accompanied by data obtained from curated real-world projects or from synthetic SE-specific datasets). \emph{RepairBench}~\cite{silva2024repairbench}, for example, contains 574 buggy Java methods and their corresponding fixed versions, which can be used to evaluate the performance of LLMs in code repair tasks. It uses the \emph{Plausible@1} metric (i.e., the probability that the first generated patch passes all test cases) and the \emph{AST Match@1} metric (i.e., the probability that the abstract syntax tree of the first generated patch matches the ground truth patch).
\emph{SWE-Bench}~\cite{DBLP:conf/iclr/JimenezYWYPPN24} is a more generic benchmark that contains 2,294 SE Python tasks extracted from GitHub pull requests.
To score the LLM's task performance, the benchmark validates whether the generated patch successfully compiles and calculates the percentage of passed test cases. Meanwhile, \emph{HumanEval}~\cite{DBLP:journals/corr/abs-2107-03374} is often used to assess code generation. 

\studytypeparagraph{Advantages}

Good benchmarks provide consistent, reproducible scores across different tasks, enabling comparison of different models (and versions). Standard benchmarks reduce research effort because each research teams uses the same evaluation methods instead of reinventing the wheel. They can foster open science practices by providing a common ground for sharing data (e.g., as part of the benchmark itself) and results (e.g., of models run against a benchmark). A body of research all using the same benchmark is easier to understand and may appear more legitimate to practitioners. In addition, benchmarks built for specific SE tasks can help identify LLM weaknesses and support their optimization and fine-tuning for such tasks.


\studytypeparagraph{Challenges}

Benchmark contamination, that is, including the benchmark in the LLM training data~\cite{DBLP:journals/corr/abs-2410-16186}, can artificially inflate LLM performance. Carefully selecting samples and creating corresponding input prompts is particularly important, as correlations between prompts may bias the benchmark results~\cite{DBLP:conf/acl/SiskaMAB24}. More fundamentally, benchmarks may lack construct validity; that is, measuring something in a reliable, standardized way does not mean one is measuring the correct something. Computer scientists do not have a strong history of rigorously assessing construct validity (cf.~\cite{ralph2024teaching}. Benchmarks usually do not capture the full complexity and diversity of software engineering work~\cite{Chandra2025benchmarks}. An LLM can simultaneously perform well on one  benchmark (e.g. \emph{HumanEval}) and poorly on another (e.g. \emph{SWE-bench}). Benchmarking may therefore lead to overconfidence and overfitting a model or tool to a specific benchmark such improvements in benchmark scores do not reflect improvements in closely-related tasks. 
Recently, \citeauthor{cao2025should}~\cite{cao2025should} proposed guidelines for  creating LLM benchmarks related to coding tasks, grounded in a systematic survey of existing benchmarks, and highlight current shortcomings related to reliability, transparency, irreproducibility, low data quality, and inadequate validation measures. For more details on benchmarks, see Section \benchmarksmetrics. 
